---
title: "Introduction to R - Exercises"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, comment = NA)
```


-----------
# Basics {.tabset .tabset-fade}

## Create a new project 

- Create a new project with the name `introduction-to-r`
    - Open RStudio by clicking on the icon or by selecting it via the start menu
    - Click on `File` - `New Project...` and select `New directory`
    - Give the directory the name `introduction-to-r` and create it as a sub-directory of your *Desktop* folder
    - Where on your hard disk is the project folder located? Try to locate it with the regular windows file explorer.
- Create a sub-directory `data` and `source` in the project's root folder.
- Download the *titanic* and the *football* dataset from the course website and place it in the `data` folder.
- In RStudio type `getwd()` into the console window. This prints out the full path of your projects root folder.
    - Type `list.files("./data")` into the console window. What's the meaning of the dot `.` in the file path?
    - What's the output of the call `list.files("../")`?

## Create a new script file (`helloWorld.R`)
- Click on `File` - `New File` - `R Script` or use the plus button in the toolbar. 
- Write `plot(rnorm(100))` into the text-editor
- Execute the selected line by pressing [Strg]-[Enter] or by clicking on the `Run` button in the toolbar. 
- What's the purpose of this command? Call the help page for the function `rnorm`. 
- *Advanced*: Write a small program which calculates the empricial mean and standard deviation for $n=10$ random draws from the standard normal distribution.
    - Use the functions `mean`, `sd` and `rnorm` 
    - E.g. type `?mean` in the console to call the help page for this function
    - What happens to the mean and deviation if $n \rightarrow \infty$? Try this in your program!
- Save the file (`File` - `Save`) in the `source` folder of your project. Give it the name `helloWorld.R`.


## KVB bikes in cologne

1. Download the script `kvb.R` from the course website and save it in the project's `source` folder.
2. Install the missing packages `XML` and `leaflet` with the function `install.packages`
    ```{r}
install.packages(c("XML", "leaflet"))
```
3. In your project in RStudio, open the script file and run the code line by line. Does the script terminate without any errors? 
4. *Advanced*
    - Change the color of the markers from red to green.
    - Try to get and visualize data for another city in which *nextbike* operates (Inside the code, you have to change the link to the nextbike api). 
    - Add a cluster effect to the map (Use the function parameter `clusterOptions = markerClusterOptions()` at the right place).
    - Change the pop-up text of the markers to display some other useful information such as the number of bikes at this place.

## Arithmetic

Perform the following calcuations using the R console:

  - $34^2 + 12^{-0.8}$
  - $\ln(4 + \sqrt{12})$
  - $\frac{2+1}{3+\pi}$
  - $\ln(e^4)$
  
*Advanced*: $\sqrt{2}^2 - 2 = 0$, right?

  - In R, why is `sqrt(2)^2 - 2` not exactly equal to zero? 
  - Do you find other examples?
  
## Function calls

- Have a look at the documentation for the function `rnorm`. 
    - How many arguments has the function?
    - What are the default arguments of the function?
    - Which of the following function calls is not valid? Why? 
    ```{r, echo = TRUE, eval = FALSE}
  rnorm(n = 10)
  rnorm(10, 2, 4)
  rnorm(mean = 2)
  rnorm(sd = 3, mean = 1, n = 3)
    ```
  - With the help of the above function calls, explain how R matches *formal* to *actual* arguments!


# Vectors and Operators {.tabset .tabset-fade}

## Basic data types

- What are the three basic data types? Write down a code example for each of them.
- What's the difference between `TRUE` and `"TRUE"`? Whats the difference between `pi` and `"pi"`?
- Use the function `as.numeric()` to cast values to a *numeric* type:
    - What's the numeric value of `TRUE` and `FALSE`?
    - Can you cast the character string `"hi"` to a numeric value? Try the same with the string `"5"`. 
    - How can you cast a numeric value to a string? To a logical value? 

## Vectors and variable assignment
- Asign the value $12$ to a variable with the name `pi`
    - Why is this dangerous?
    - Remove the variable by calling the `rm()` function with the appropriate argument.
    
- Use the `c()` function to store the results from the calculations in exercise *1.4* in a vector . Assign the vector to a variable with the name `values`.

- Try to find the name of the built-in functions and answer the following questions:
    - What's the sum and the product of the elements in the vector?
    - What is the smallest/ largest value?
    - How can you sort the vecor?
    - How can you determine the length of the vector? 

- Call the `ls()` function to list all user variables in the global envrionment.
    - How many variables are listed? Try to answer this question programmatically. 
    - What is the data type of the object returned by a call of `ls()`?
    

    
## `descriptive.R`
- Open a new script file and save it in the `source` folder under the name `descriptive.R` 
- Write a program which calculates the mean, the standard deviation, the range, the maximum and the minimum of an arbitrary numeric vector `x`
    - You program should start with the following lines of code:
    ```{r}
# define some arbitrary numeric vector with varying length e.g.
x <- c(1, 4, 2, 9)
# calculate the statistics
```
    - Use only the built-in functions `max()`, `min()`, `length()` and `sum()` 
    - The range is defined as $x_{max} - x_{min}$

## Vector generation
- Try to generate the following sequences with the help of the functions `c()`, `rep()` and `seq()`:
    - $(1, 2, \dots, 100)$
    - $(-10, -9.5, -9, -8.5, \dots, -4.5, -4)$
    - $(TRUE, FALSE, TRUE, FALSE, TRUE, FALSE)$
    - $("A", "B", "C", "A", "B", "C", "A", "B", "C")$
    - $(100, 90, 80, \cdots, 10, 0, -10)$
    
- *Advanced* (Tip: Have a look at the documentation for the function `rep()`):
    - $(1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, \cdots, 100, 100, 100)$
    - $(1, \dots, 100, 1, \dots, 100, 1, \dots, 100, 1, \dots, 100)$
    - $(1000000, 100000, 10000, 1000, 100, 10, 1)$
    
## Casting 
Have a look at the two vectors: 
```{r, eval = F}
x <- c("TRUE", "FALSE", "TRUE", "TRUE")
y <- c("2", "4", "5")
```
Execute the following expressions in R and explain the result.
```{r, eval = F}
as.numeric(as.logical(x))
is.numeric(y)
is.character(x)
is.logical(as.character(TRUE))
```

Try to guess the result of the following expressions without using a computer
```{r, eval = F}
is.logical(0)
is.logical(as.logical(0))
is.numeric(as.logical(x))
as.character(1243)
```

## Logical expressions
Have a look at the two values `x` and `y` and the vector `a` and `b`
```{r, eval=F}
x <- 9^2
y <- "test"
a <- c(4, 6, 5, 7, 10, 9)
b <- c(4, 6, 2, 7, 2, 3)
```

Execute the following expressions in R and try to explain the result
```{r}
(4 < 5) & (sqrt(x) == 9)
y == "Test" | y == "test"
x < 100 | y != "hallo"

a > 5 & b < 7
all(a > 4 | b < 6)
any(a < 2)
length(a) == length(b)
```

Try to guess the result of the following expressions without using a computer
```{r, eval = F}
a < 4 & b > a
a == length(b)
any(sqrt(x) == a)
all(a < b)
length(x)
```


## Logical expressions and vectorization
- Create a vector of logical values and apply the `sum()` and `mean()` function to it. What do you observe?
- With the function `rnorm()`, sample $n = 100$ values from the normal distribution. Count the *relative* and *absolute* number of observations which are
    - positive
    - between $-1$ and $1$
    - larger than two times the empirical standard deviation of the sample
- *Advanced*: Given the previous three expressions, calculate the *analytical* probability values with the function `pnorm()` for some random normal variable with $X \sim N(\mu_x,\sigma_x)$. E.g. calculate the probabilities:
    - $P(X > 0)$
    - $P(-1 \leq X \leq 1)$
    - $P(X > 2\sigma_x)$
    

## Missing values
- What is the keyword for a missing value? Create a vector with numeric values and some missing values. 
- Use the function `is.na()` on your vector to test for missing values. What is the return type of the function `is.na()`? 
- Use the `sum()` function together with `is.na()` to count the number of missing values in the vector. 
- Have a look at your code in `descriptive.R`. 
    - What happens to the code, if the vector `x` contains missing values? 
    - Try to rewrite your code such that missing values are handled (E.g. look at the further function arguments of the functions `mean()`, `sum()`, ...).
    - Expand the code by an expression, which counts the number of missing values. 
    



# Data import and data frames {.tabset .tabset-fade}

## `.csv` data
Open the titanic dataset with a text editor (e.g. right click on the file - `edit with notepad++`)

  - How many variables/columns has the dataset?
  - How many passengers are included in the dataset?
  - Does the file contain a header line?
  - How are the values separated?
  - What's the encoding of the file?

## The titanic dataset
- Open a new script file, save it in `source` and give it the name `titanic.R`.
- Use the `read.csv()` function with the correct file path, to import the titanic dataset into R (Remember to also include the argument `stringsAsFactors = FALSE`).
- Have a look at each column and determine it's datatype. You can also use the `summary()` or the `head()` function for this. 
- Try to answer the following questions `programmatically`:
    - How many passengers and which variables are included in the dataset? 
    - How many people died on the titanic? How many people survived? Calculate the overal survival rate (E.g. the number of survivors as a fraction of the overall number of passengers).
    - How many of the passengers are women?
    - How many missing values does the variable `age` contain? 
    - What's the age of the youngest and oldest passenger on the titanic?
    - How many passengers are in the first class?

## Frequency tables
With the titanic dataset and the function `table()` answer the following questions:

  - How many men are in the third class?
  - How many people died in the first class? How many in the third class?

In addition, use the functions `prop.table()` to transform the tables into relative frequency tables:

  - Calculate the survival rate separately for each class. Did people from the first class had a higher chance to survive?
  - Did women have a higher chance to survive?

## Variable creation and updating
- Add a new column `isMale` to the dataset which is `TRUE` if the passenger is male. 
- Change the data-type of the `survived` column from `numeric` to `logical`. 
- Add a new column `isChild` to the dataset which is `TRUE` if `age < 18` and `FALSE` else. Are missing values preserved?
- *Advanced*: Add a new column which is `TRUE` if the `name` contains the string `"Miss"` and `FALSE` else.
    - Use the function `grepl()`
    - Did unmarried woman had a greater chance to survive?

## *Advanced*: Significant differences in the survival rates (chi sqared test)
Use the function `chisq.test` to check if there is a significant relationship between the class and the survival rate of a personen.

  - Use the `table` function to create a contingency table. 
  - Whats the null hypothesis of the test? 
  - How large is the p-value and what can you conclude?
  
## *Advanced*: The football dataset

- Import the football dataset using the `read.csv2()` function. Why do you have to use `read.csv2()` instead of `read.csv()`? 
- How many games are in the dataset?
- How many unique countries are included in the dataset? 
- How many games from the `FIFA World Cup` are included in the dataset? 
- Think of some questions you want to answer with the dataset and if you can allready answer them with the tools you have learned so far.
  


# Functions and control structures {.tabset .tabset-fade}

## Function `countMissings()`
Write a function `countMissings` which accepts a numeric vector `x` as input argument and returns the number of missing values `NA` in the vector. 

  - Use this template as a start: 
    ```{r, eval = F}
countMissings <- function(x){
  # calculate nMissings here

  # return the number of missing values
  return(nMissings) 
}
```
  - Apply the function to the `age` variable from the `titanic` dataset. 
  - *Advanced*: If the vector contains *only* missing values, a warning should be printed to the console. Use the function `warning()` and an `if` construct for this. 

## Turn `descriptive.R` into a function
Wrap your code from `descriptive.R` into a function with the name `descriptive()`. The function accepts a numeric vector `x` as input argument, calculates the descriptive statistics and returns them in a *named* vector. Use this template as a start: 
```{r, eval = F}
descriptive <- function(x){
  # calculate the descriptive statistics for x here
  
  # return the result
  return(result) # result should be a named vector containing the descritpive statistics
}
```

  - Tip: You can create a named vector like this: 
    ```{r, echo = TRUE, eval = FALSE}
result <- c(mean = 0.4, sd = 4, range = 2)
```
  - Apply the function `descriptive()` to the `survived` and `pclass` column from the `titanic` dataset. 
  - *Advanced*: What happens if you apply your function to the `age` variable which contains missing values. Think of ways to correctly handle missing values (e.g. you could add an additional argument `removeNA` to your function)
  - *Advanced*: If the type of the vector `x` is not `numeric`, the function should terminate with an error (look at the function `stop()`). 



## The `sapply()` function
- use `sapply()` with the `countMissings()` function from exercise $4.1$ to count the number of missing values in each column of the `titanic` dataset.


## Visibility of variables
What's wrong with the following lines of code? Rewrite the function `mySum` such that it can be used without errors. 
```{r}
mySum <- function(a){# calculates the sum of two numbers a and b
  return(a+b)
}
mySum(4)
b <- 6
mySum(2)
```
Look at the following code:
```{r}
norm <- function(x){# calculates the eucledian norm of a vector
  xNorm <- sqrt(sum(x^2))
  return(xNorm)
}
x <- c(1, 2, 1)
x <- norm(x)
xNorm
x
```
  - Does the code terminate without any errors? Where is the mistake? Correct the mistake.
  - In this example, what's the *formal* and the *actucal* argument of the function `norm()`?
  - What's the value of the variable `x` after the code terminates?
    

## *Advanced*: Function `normalize()`
- Write a function `normalize()` which standardizes a numeric vector `x` such that
$$ 
x_{norm} = \frac{x-\bar{x}}{x_{sd}}
$$
    - If the vector contains missing values a warning via the function `warnings()` should be printed to the console. 
    - If the vector is not `numeric` (e.g. it contains logical values), the function should be terminating with an error. Use the function `stop()` and `is.numeric()`.
- Rewrite the function `normalize` such that the user can pass two parameters `m`and `s` with
$$ 
x_{norm} = \frac{x-m}{s}
$$
    - Think of *downward compability*: A user should still be able to call the function via `normalize(x)`. You can achieve this by using *default arguments*. 


# Extraction and updating {.tabset .tabset-fade}

## Vector subsetting
- Which methods for extracting values from a vector do you know?
- Sample $n = 100$ values from the standard normal distribution and subset the sample by selecting
    - only positive values
    - values between $-1$ and $1$ 
    - the first and the last value
    - everything but the first value

## Test
Look at the two vectors:
```{r, eval = F}
x <- c(2, 3, 4, 4)
y <- c(TRUE, FALSE, FALSE, TRUE)
```
Without using a computer: What's the result of the following expressions?
```{r, eval = F}
sum(x[y])
sum(x[!y])
sum(y[x])
sum(y[-x])
# advanced
which(x > 3)
which(x[y] > 3)
```

## Remove missing values from a vector
How can you remove missing values from a vector? Write a function `removeNA()` which takes some numeric vector `x` as input and returns the vector with missing values removed.

  - Find an expression for selecting only non-missing values from the vector `x` (e.g. use the function `is.na` together with extraction method `[]`)
  - Inside the function create a variable `xClean` containing the subset of `x` with the missing values removed
  - Return the vector `xClean`
  
A call of this function should look like this: 
```{r, echo = FALSE, eval=TRUE}
removeNA <- function(x){
  x[!is.na(x)]
}  
```

```{r, echo = TRUE, eval = TRUE}
x <- c(2, 3, NA, 4)
removeNA(x)
```

    
## Set negative numbers to zero
Write a function `negToZero()` which sets negative numbers in a vector `x` to $0$. A call of this function should look like this: 
```{r, echo = FALSE, eval=TRUE}
negToZero <- function(x){
  x[x < 0] <- 0
  x
}  
```

```{r, echo = TRUE, eval = TRUE}
x <- c(-10, 3, -2, 3, 4)
negToZero(x)
```


## *Advanced*: Recoding 
Write a function `codeToNA()` which converts some user defined numbers in a vector `x` into `NA`s. 

  - Tip: Use the `%in%` operator to update the values inside the vector `x`. 
  - To understand how the `%in%` operator works, have a look at this example: `1:10 %in% c(1,3,5,9)`
  
A call of this function looks like this:
```{r, echo = FALSE, eval=TRUE}
codeToNA<- function(x, a){
  x[x %in% a] <- NA
  return(x)
}
```

```{r, echo = TRUE, eval = TRUE}
x <- c(2, 3, -99, -999, NA, 4)
codeToNA(x, c(-99, -999))
```

## Outlier detection

Write a function `removeOutliers()` which identifies positive outliers in a numeric vector `x`.

  - Here, a positive outlier $x_i$ is defined as $x_i: x_i > \bar{x} + 3*sd(x)$. 
  - The function should return the vector without outliers. 
  - *Advanced*: Enhance the functionality by allowing the user to specify if only *positive*, only *negative* or both *positive* and *negative* outliers should be removed. 

A call of this function looks like this:
```{r, echo = FALSE, eval=TRUE}
removeOutliers<- function(x){
  isOutlier <- x > (mean(x) + 2*sd(x))
  x[!isOutlier]
}
```

```{r, echo = TRUE, eval = TRUE}
x <- c(2, 3, 120,1.5, 2, 1, 0, 999)
removeOutliers(x)
```

- *Advanced*: You can also write a function `detectOutliers` which gives the *indices* of the outliers in a vector. Are there any outliers in the age column of the titanic dataset?  

## Extraction and updating of data frames

- Add a new column `ageImputed` to the dataset which contains the age of the passenger but with `NA`s replaced by the overall average age (see [Mean imputation](https://www.iriseekhout.com/missing-data/missing-data-methods/imputation-methods/)).
- Create a copy of the titanic dataset for which the passengers (e.g. the rows) with a missing value are removed.
- Filter the titanic dataset and select only: 
    - the survivors
    - women in the first class
    - male children.
    
## *Advanced*: Two sample t-tests
- What is the average age for passengers which have a `Miss` in their name (comare with exercise 3.4)?
- Use the function `t.test` to perform a two sample t-test and check wether the difference in the mean age between Misses and all other passengers is significant. 
- What's the meaning of the confidence interval? 
- Use the same test but now just use a one-sided test (e.g. test if the difference in the means is less than zero).

## Ordering of data frames
- Order the rows of the titanic dataset such that the age is in *decreasing* order.
- What's the name and age of the youngest/ the oldest passenger? Did he or she survive?
- *Advanced*: Select the top $10$ youngest passengers from each class. How many of them survived?


# Visualizations {.tabset .tabset-fade}

## Step by step visualization
- Complete the step by step visualization from the slides
- Add additional elements or labels
- Change the colors, the point size or the line type
- Save your plot as a pdf document

## Visualization examples
- Choose a visualization from the examples on the slides and complete the plot, e.g.

  - add a title, subtitle, caption
  - change the theme
  - ...

## Visualization project

- Form groups of two to four and choose a dataset
- Develope a small research question or replicate a question from the exercises
- In a new .R script file read in the data, clean and transform it
- If needed, apply a regression model or conduct a statistical test
- Create 1 or 2 visualizations and upload the code to ILIAS
    - Make sure that your script file contains no errors, is readable and sharable



# Advanced exercises {.tabset .tabset-fade}

## Matrics
Use the `matrix` function and create the following matrics
$$
A=
\begin{pmatrix}
2 & 4 \\
1 & 6 \\
\end{pmatrix}
, B =
\begin{pmatrix}
1 & 3 & 9\\
8 & 2 & 4\\
\end{pmatrix}
, C =
\begin{pmatrix}
TRUE & TRUE & FALSE\\
TRUE & FALSE & TRUE\\
\end{pmatrix}
$$
Calculate 

  - $A*B$
  - $B'*A$
  - $(A*B*A')^{-1}$.

## Two-sample t-test for equal means
- Write a function `myTTest` which tests, if the mean of two given numeric vectors `x` and `y` is statistically different from one another (see [here](https://en.wikipedia.org/wiki/Student%27s_t-test#Independent_two-sample_t-test) for a detailed explanation)
    - The input arguments of the function are the vectors `x`, `y` and a significance level `alpha`
    - The null hypothesis is: $H_0: \mu_x = \mu_y$.
    - The test value can be calculated as:
      $$
      T = \frac{\bar{x}-\bar{y}}{\sqrt{\frac{1}{n}x_{sd}^2+ \frac{1}{m}y_{sd}^2}}
      $$
    - $H_0$ is rejected if $|T| > u_{1-\frac{\alpha}{2}}$
    - $n$, $m$ are the length of the vectors $x$ and $y$
    - $u_{1-\frac{\alpha}{2}}$ is the quartile of the normal distribution (see `qnorm()`)
    - The function should return a list with the mean of each of the two vectors, the test value and a logical value indicating the test result
    
- Compare your results with the built in function `t.test`
- *Advanced*
    - Calculate p-values or produce a nice output (e.g. with stars indicating the strength of significance)
    - If the sample size $n,m<40$ a warning should be printed (see `warning()`)

## Write your own linear regression function
Write a function `myRegression` which performs a multivariate linear regression analysis given some outcome `y` and some design matrix `X`. The model is given by the equation:
$$
y = X\beta + \epsilon,~ \epsilon \sim N(0, \sigma^2)
$$
The function should calculate the following statistics:

  - The regression coefficients: $\hat{\beta} = (X'X)^{-1}X'y$
  - The number of observations $N$ and the number of regressors $K$
  - The residuals: $e = y - X\hat{\beta}$
  - The error variance $\hat{\sigma^2} = \frac{\sum_{i = 1}^{N}{e_i^2}}{N-K}$
  - The coefficient of determination $R^2 = 1 - \frac{\sum_{i = 1}^{N}{e_i^2}}{\sum_{i = 1}^{N}{(y_i - \bar{y})^2}}$

The results should be returned using a named list. Check your calculations using the built-in function `lm`.

*Advanced*: Enhance the function with a t-test for the coefficients and calculate p-values and confidence intervals

*Tip*: R can do matrix arithmetic such as addition, multiplication and inversion: Have a look at `help("%*%")` and `?solve`


## Modelling survival rates with linear regression

Use the titanic dataset and the examples from to slides:

  - Run a simple linear regression model with the formula `survived ~ ageImputed`. Is the age significant (`ageImputed` was calculated in exercise 5.7)? 
  - In addition, include the `pclass` and the `sex` as independent variables. Is the age now significant? 
  - In the model above, what's the effect of the age on the survival rate (E.g. how much does the probability of survival decreases for passenger who is one year older)?
  - Include the `embarked` variable. Are there any significant effects?
  - Use the `predict.lm` function to predict the probability of survival for each passenger. Compare the predicitons to the real outcome. How good is the fit?
  - *Advanced*: What's the probability of survival for a $20$ year old male in the third class? Use the `predict.lm()` function with the `newdata` argument. 
  


  
```{r, echo = FALSE, eval = FALSE}
m <- lm(age ~ pclass + sex + embarked, data = titanic)
pred <- predict(m)
plot(pred, titanic$age[!is.na(titanic$age)])
```


  



## String operations
Try to squeeze out some information from the `name` column in the `titanic` dataset: 

  - Look for family names and - together with the columns `embarked` and `pclass` - identify family members.
  - Extract the title (e.g. *Mr*, *Miss*, *Master*, ...) from the name.
  - What else can you do? 
    
Tip: Install the `stringr` package, load it with `library(stringr)` and have a look at the various functions having the naming convention: `str_<action>()` (e.g. `str_replace()`, `str_detect()`). 



## Logistic regression modelling and visualization
Use the `ggplot2` package, the `glm()` function and the `titanic` dataset to model and visualize survival rates on the titanic. Use a logistic regression model with the independent variables *age*, *pclass* and *sex* and the dependent variable *survived*. 

The final figure could look like this (A script to produce this figures can be found in the [scripts](./scripts.html) section):

```{r titanic-survival-rates, echo = FALSE, eval = TRUE, dpi=300, fig.width=8}
library(ggplot2)
titanic <- read.csv("./www/titanic.csv", stringsAsFactors = FALSE)
titanic$pclass <- as.factor(titanic$pclass)
titanic$sex <- as.factor(titanic$sex)

form <- survived ~ pclass + sex + age + pclass:sex + pclass:age + sex:age
m <- glm(form, data = titanic, family = "binomial")

dat <- expand.grid(sex = c("male", "female"), pclass = as.factor(c(1, 2, 3)), age = 0:70)
pred <- predict(m, newdata = dat, type = "link", se.fit = TRUE)
dat$survived <- plogis(pred$fit)
dat$upper <- plogis(pred$fit + 1.96*pred$se.fit)
dat$lower <- plogis(pred$fit - 1.96*pred$se.fit)

ggplot(dat, aes(x = age, y = survived, color = sex, fill = sex, ymin = lower, ymax = upper)) + 
  geom_line() + 
  geom_ribbon(linetype = 0, alpha = 0.1) + 
  facet_grid(~ pclass, labeller = as_labeller(c("1" = "first class", "2" = "second class", "3" = "third class"))) + 
  ggtitle("Survival rates on the titanic", 
          paste0("Logistic regression model equation: ", format(form), "\n95% confidence intervals, N: ", m$df.null + 1)) + 
  theme(legend.position = "top", legend.title=element_blank(), legend.justification = "left")


```

A guidance on how to proceed:

  - Import the `titanic` dataset and use the function `as.factor()` to transform the *pclass* and the *sex* variable into the `factor` datatype.
  - Use the `glm` function with the formula: `survived ~ pclass + sex + age + pclass:sex + pclass:age + sex:age` and the `family = "binomial"` argument.
      - with the `:` in the formula, interaction effects can be specified
      - with `family = "binomial"` one can specify a logistic regression model
  - Apply the `summary()` function to the model object and try to interpretate the coefficients. Why is it important to transform the *pclass* variable to the `factor` datatype befor applying the model?
  - Use the model, to predict survival rates over the possible range of `age`, `sex` and `pclass`
      - with the `expand.grid()` function, create a new `data.frame` filled with "artifical" passenger data: for each possible outcome of `sex` and `pclass` you should vary the `age` from $0-70$. 
      - It is important that the new data has the same columns names as the variable names in the model (e.g. `sex`, `pclass` and `age`).  
      - use the function `predict()` on the model object and the new data to predict the survival rates.
      - add a new column with the predictions to the `data.frame` containing the new passenger data.
  - Use the `ggplot()` function to plot the passenger data
      - Define the asthetics inside `ggplot()` with the `aes()` function. Use the asthetics `x`, `y` and `color`. 
      - Add the trend lines with `geom_line()`
      - Use `facet_grid(~pclass)` to create three separate plots
      - Add a title with `ggtitle()`
  - *Further steps*: Find out how to add confidence bars to the plot
      - The `predict()` function can return standard errors for the predictions
      - The `geom_ribbon()` function adds an interval arround the trend lines. Use the additional asthetics `ymin` for the lower confidence bound, `ymax` for the upper confidence bound and `fill` for the transparent fill color. 

  
