---
title: "Introduction to R: Vectors and Operators"
author: "Malte Bonart"
output: revealjs::revealjs_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
# http://directpoll.com/c?XDVhEtVHRZBEc4Ln7AxJqIT3LugvOjDt
```

# Vectors

## Introduction
- `vector`: basic data structure in R
- one dimensional array of the same data type
- three data types: `numeric`, `character`, `logical` 
- `c()` function to combine elements to a vector

## Numeric, character, logical 
- `character`: elements are specified in quotes: `"hello"`
- `logical`: keywords `TRUE` or `FALSE`
- `numeric`: regular numbers

## Numeric, character, logical 
```{r}
c(3+4, 2^3, sqrt(123), exp(2))
c(TRUE, FALSE, FALSE)
c("hi", "by")
```

## Length of a vector
- vectors always have a defined length
```{r, eval = TRUE, echo = TRUE}
m <- seq(1, 12)
length(m)
```

## Creation
- the `c()` function: `c(1, 2, 12, 3)`
- the `rep()` function: `rep("A", times = 4)`
- the `seq()` function `seq(from = 1, to = 4, by = 0.25)`

## Sequences
- `c()` to combine values to a vector of the same data type. 
- `rep()` to repeat the values of a given vector
- `seq()` to generate sequences with fixed length
- `x:y` short for `seq(x, y, by = 1)`

## Casting and testing
- using the functions `as.numeric()`, `is.numeric()`, `as.character()`, `is.character()`, ... one can test and cast an object from one type to another
```{r}
as.numeric("4")
is.logical(0)
is.logical(TRUE)
```

## Names of a vector
```{r}
month <- 1:12
month
names(month) <- c("Jan", "Feb", "Mär", "Apr", "Mai", "Jun", "Jul", "Aug", "Sep", "Okt", "Nov", "Dez")
month
```

## Random numbers
- `rnorm()` random numbers from the normal distribution
- `rbinom()` - random numbers from the binomial distribution
- `runif()` - uniform distribution 
- `rt()` - t distribution
- `sample()` - sample values with or without replacement from a vector

# Operators
## Overview
- Numeric operators: $numeric \times numeric = numeric$
- Relational operators: $numeric \times numeric = logical$
- Logical operators:  $logical \times logical = logical$
- Assignment Operator `<-`: $name \times value = object$

## Numeric operators: $numeric \times numeric = numeric$
```{r}
3^2
-3+5
3*7
9/3
```
## Relational operators: $numeric \times numeric = logical$
```{r}
4 < 5
3 != 4
5 == 2
"a" < "b" 
"hallo" == 10
```
## Logical operators: $logical \times logical = logical$
```{r}
TRUE | FALSE
TRUE & FALSE
!TRUE
```
## Assignment Operator $name \times value = object$
```{r}
y <- 10
x <- 2
x
```

## Vectorization
- Most operators and many functions work for vectors 
- This is usefull for fast calculations and checking on large vectors
```{r}
numbers <- c(1, 2, 5, -100, 23, 43)
sqrt(numbers)
numbers < 0
```


# Missing values
##  Missing values
- very common when working with real datasets
- Keyword: `NA`

## Examples
```{r}
val <- c(3, 5, 2, NA)
sum(val)
sum(val, na.rm = T)
is.na(val)
sum(is.na(val))

```




