---
title: "Introduction to R: Data import and data frames"
author: "Malte Bonart"
output: revealjs::revealjs_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
# http://directpoll.com/c?XDVhEtVHRZBEc4Ln7AxJqIT3LugvOjDt
```

# Data import

## Table data

- most of the available data is structured in a *table* representation
    - rows: observations, cases, subjects
    - columns: variables, features
- table data can be stored 
    - in plain text: `.txt`, `.csv`, `.tsv`, ...
    - in a binary format: Excel, Stata, SPSS, ...
    - in a relational database: `SQL`, ...

## Text table data
- central function is `read.table()` to read table data from the hard drive
- questions to answer before importing table data:
    - has the data a header line?
    - how are values separated (e.g. `","` or `";"`)
    - what's the encoding (e.g. `UTF-8` or `latin-1`)
- use the argument `stringsAsFactors = FALSE` 

## Binary table data
- there exists various packages for importing binary table data
    - `haven` package for Stata, SAS and SPSS files
    - `readxl` package for Excel files
- usually one can always export the data to text format and then import it using `read.table`

## Other data formats

- Web APIs often represent data in `json` or `xml` (text-)format
    - packages `XML` and `jsonlite`
- R uses the binary `.RData` or `.rds` format to save objects persistently
- Spatial data is often stored in a *GIS* file format (shapefile)
    - packages `maptools`
    
    
# Data frames

## Introduction
- when importing table data, the natural R representation is a `data.frame`
- two-dimensional data structure with rows and columns
- each column is a `vector` of some data-type (e.g. `numeric`, `logical`, `character`, ...)

## Example: creation
```{r}
myDF <- data.frame(1:10, rep(TRUE, 10), letters[1:10])
myDF
```

## Example: data import
```{r}
titanic <- read.csv("./www/titanic.csv", stringsAsFactors = FALSE)
titanic <- read.table("./www/titanic.csv", header = TRUE, 
                      sep = ",", stringsAsFactors = FALSE)
head(titanic)
```

## Example: `summary` function
```{r}
summary(titanic)
```

## Length
- use the functions `nrows()`, `ncols()`
- `length()` also gives the number of columns
```{r}
nrow(titanic)
ncol(titanic)
length(titanic)
```

## Names
- function `colnames()` or `names()` for the column names
- function `rownames()` for the row names (used less frequently)
```{r}
colnames(titanic)
```

# Extraction and adding of variables

## Column extraction
- each column of a `data.frame` is a vector of some data type
- columns can be extracted using the `$` sign
```{r}
class <- titanic$pclass
class
```

## Column creation
```{r}
titanic$isOld <- titanic$age > 60
titanic$isOld
```

## Overriding of existing columns
```{r}
titanic$age <- 12
head(titanic$age)
```

## Deletion of columns
```{r}
titanic$age <- NULL
head(titanic)
```


# Frequency and contingency tables

## Introduction
- used to represent the distribution of one or two *categorical* variables
- total frequencies
- relative frequencies (row, column, or absolute percentage)

## Example total frequencies
```{r}
table(titanic$pclass)
table(titanic$sex, titanic$pclass)
```

## Example relative frequencies, row percentage
```{r}
tab <- table(titanic$sex, titanic$pclass)
tab
prop.table(tab, margin = 1)
```

## Example relative frequencies, column percentage
```{r}
tab <- table(titanic$sex, titanic$pclass)
tab
prop.table(tab, margin = 2)
```

## Example relative frequencies, absolute percentage
```{r}
tab <- table(titanic$sex, titanic$pclass)
tab
prop.table(tab, margin = NULL)
```
