---
title: "Introduction to R: Functions and control structures"
author: "Malte Bonart"
output: revealjs::revealjs_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
titanic <- read.csv("./www/titanic.csv", stringsAsFactors = FALSE)
```

# `if`- `else` conditions

## Introduction
`if(cond) {true.expr} else {false.expr}`  

- Binary case switch: If `cond` is `TRUE` then run `{true.expr}` else run `{false.expr}`. 
- Applications
    - control random events
    - call special handlers in case of missing values, wrong data type, ...
    - inside functions

## Example
```{r}
if (4 < 6){
  val <- 4
} else {
  val <- 4 + 2
}
val
```


# Functions
## Introduction
- data structures represent information, functions 'do' something with information
- practically no difference between user-defined and built-in functions
- cleaner and less error-prone code
- should be defined at the *beginning* of a script file or in an extra file

## Example
```{r, eval = TRUE}
hello <- function(name){
  words <- paste("Hello", name)
  return(words)
}
hello("Lisa")
hello("Ben")
```

## Definition
- Keyword: `function`
- Formal function arguments: `(arg1, arg2, ...)`
- Function body: `{...}`
- A `return` statement

## Example

```{r}
mySum <- function(a, b){
  sum <- a + b
  return(sum)
}
mySum(a = 4, b = 10)
mySum(a = 4, b = NA)

```

# Visibility of variables
## Input and output
- When passing objects to a function a *copy* of this object is created
- The copy only exists inside the function. The global object is not changed.
- The return value of a function can be saved as a new variable

## Example
```{r}
addTwo <- function(x) {
  x <- x + 2
  return(x)
}
x <- 6
addTwo(x)
x                 # x is still as before
x <- addTwo(x)    # the result of the function is assigned to the variable x
x                 # now x has changed
```

## Variables inside functions
Variables which are declared inside a function are not accessible from outside the function
```{r, error = T}
f <- function(x, y) {
  z <- x + y
  return(x)
}
f(2, 3)
z
```

## Functions and the global environment I 
Variables which were defined outside of a function are also visible inside the function
```{r}
y <- 5
h <- function(x) {
  return(x*y)
}
h(2)
```

## Functions and the global environment II
But what happens, if the object `y` is removed from the global environment?
```{r, error = T}
rm(y)
h(2)
```
**Never use global variables inside a function!**


# The `sapply` function

## Introduction

- sapply a function to each column of a `data.frame`
- *functional* programming: pass a function as function argument to another function
- family members: `lapply`, `sapply`, `vapply`, ...

## Example: Check for missing values, the bad way
```{r, eval = TRUE, echo = FALSE}
countNA <- function(x){
  sum(is.na(x))
}
```


```{r}
countNA(titanic$pclass)
countNA(titanic$sex)
countNA(titanic$age)
countNA(titanic$name)
```

## Example: Check for missing values, the `sapply` way
```{r}
sapply(titanic, countNA)
```

## Advantages

- avoid repetitions
- cleaner code and less errors
- becomes powerful with user defined functions






