---
title: "Introduction to R: Extraction and updating of data-strucures"
author: "Malte Bonart"
output: revealjs::revealjs_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
# http://directpoll.com/c?XDVhEtVHRZBEc4Ln7AxJqIT3LugvOjDt
```


## Overview of important data strucutures
```{r, echo = F, results = 'asis'}
tab <- data.frame(Homogenous = c("vector: c()", "matrix: matrix()"), Heterogenous = c("list: list()", "data.frame: data.frame()"))
rownames(tab) <- c("One dimension", "Two dimensions")
knitr::kable(tab, format = "html")
```

# Recap: vectors
## Basics
- data-types 
    - `numeric`
    - `character`
    - `logical`
- missing values: `NA`
- names: `names()`
- length: `length()`
- many functions and operators can be applied to single values *and* vectors

## Creation `c()`
```{r}
x <- c(A = 4, B = 12, C = 34)
x
names(x)
length(x)
is.numeric(x)
```

## Test
What's the result of the following expressions?
```{r, eval = F}
x <- c(A = 1, B = 5, C = 2)

length(x)             # 1

is.numeric(names(x))  # 2

length(x == 1)        # 3

sum(x == 1)           # 4

is.logical(x != 2)    # 5
```
---
```{r}
x <- c(A = 1, B = 5, C = 2)

length(x)             # 1
is.numeric(names(x))  # 2
```
---
```{r}
x <- c(A = 1, B = 5, C = 2)

length(x == 1)        # 3
sum(x == 1)           # 4
is.logical(x != 2)    # 5
```


# Extracion and updating

## Extraction with `x[y]`
4 possible methods to extract values from a vector `x` using another vector `y`:

  - `y` is a `numeric` vector with positive indices from `1` to `length(x)`
  - `y` is a `logical` vector of size `length(x)` 
  - `y` is a `numeric` vector with negative (exclusive) indices from `-1` to `-length(x)`
  - (`y` is a `character` vector of names corresponding to `names(x)`)
  
## Example
```{r}
x <- c(A = 1, B = 4, C = 6, D = 2)
x
```

---

- `y` is a `numeric` vector with positive indices from `1` to `length(x)`
```{r}
x[c(1, 2)]
x[1:length(x)]
x[c(1, 1, 1)]
```

---

- `y` is a `numeric` vector with negative (exclusive) indices from `-1` to `-length(x)`
```{r}
x[c(-1, -2)]
x[-1]
x[-length(x)]
```

---

- `y` is a `logical` vector of size `length(x)` 
```{r}
x[c(TRUE, FALSE, FALSE, TRUE)]
x[x > 3]
x[x == 1 | x == 4]
```

---

- `y` is a `character` vector of names corresponding to `names(x)`
```{r}
x[c("B", "C")]
x["D"]
```


## Update values with `x[y] <- z`
Extracted values can be directly updated (in-place).
```{r}
x <- c(1, 4, 2, NA, 9, NA, NA, -99)
x
x[x == -99] <- NA
x
x[is.na(x)] <- -1
x
```

## Replace vectors with `x <- x[y]`
Extracted values can be assigned to a new variable or override the original variable 
```{r}
x <- c(1, 4, 2, NA, 9, NA, NA, NA)
x
x <- x[!is.na(x)]
x
```


# Extraction and updating for data frames

## Introduction
- data frames are *two-dimensional*
- each row represents a case, an observation, a person, ...
- each column represents an outcome, a variable, ...
- data frames have column names `names()`

## Extraction: DF[x, y], DF[x, ], DF[, y]

- `x` chooses the rows, `y` chooses the columns: `DF[x, y]`
- To select only rows/ columns
    - `DF[x, ]`: select only rows
    - `DF[, y]`: select only columns
- Extraction methods are analogous to vectors
- Special case: `DF$y` to select only the column with the name `y`

## Extraction methods for rows and columns

- a vector of positive indices: `DF[c(1, 2), ]`
- a vector of negative indices: `DF[-1, -2]`
- a logial vector: `DF[x > 2, ]`
- a vector with columnnames: `DF[, "pclass"]`

## Example I

```{r}
titanic <- read.csv("./www/titanic.csv", stringsAsFactors = FALSE)

titanic[c(1, 2, 3), "age"]
titanic[c(1, 2, 3), 1:3]
```

---
```{r}
titanic[1, ]
titanic$name[1:3]
```


## Example II


```{r}
titanic[titanic$age > 10 & titanic$age < 20 & !is.na(titanic$age), ]
```



---

```{r}
titanic[titanic$sex == "male", ]
```

---

```{r}
rowFilter <- titanic$age > 75 & !is.na(titanic$age)
colFilter <- c("name", "survived", "pclass")
titanic[rowFilter, colFilter]
```

## Update values `DF$x[y] <- z`
- values can be updated in place
- usually the `$` - sign is used to select the column first 

## Example
```{r}
titanic$sex[1:10] # convert the column to 0, 1 values
titanic$sex[titanic$sex == "male"] <- 0
titanic$sex[titanic$sex == "female"] <- 1
titanic$sex[1:10]
titanic$sex <- as.numeric(titanic$sex)
titanic$sex[1:10]
```

---

```{r}
sum(is.na(titanic$age)) # age contains missing values
meanAge <- mean(titanic$age, na.rm = TRUE) # caculate the average age 
meanAge 
titanic$age[is.na(titanic$age)] <- meanAge # impute missing values with the average age
sum(is.na(titanic$age))
```


## Delete Rows or Columns
as with vectors, data frames can be replaced by a subset 
```{r}
# only female
titanic <- titanic[titanic$sex == 1, ]
# delete the third column
titanic <- titanic[, -3]
# another possibility
titanic[,3] <- NULL

head(titanic)
```


## The `which` function
```{r, eval = TRUE, echo = FALSE}
titanic <- read.csv("./www/titanic.csv", stringsAsFactors = FALSE)
```

- Returns the indices at which a `logical` vector is `TRUE
```{r}
x <- c(FALSE, TRUE, TRUE, FALSE)
which(x)
which(titanic$survived == 1 & titanic$pclass == 1)
length(which(x))
sum(x)
```
---
```{r}
x <- c(FALSE, TRUE, TRUE, FALSE)
length(which(x))
sum(x)
```


# Ordering by variables and selection of maxima and minima

## Ordering of a dataset
```{r}
titanic <- titanic[order(titanic$age), ]
head(titanic)
```

## Select the top/ bottom 5
```{r}
youngest <- order(titanic$age)[1:5]
oldest <- order(titanic$age, decreasing = TRUE)[1:5]
titanic[youngest, -3]
titanic[oldest, -3]
```
---
```{r}
oldest <- order(titanic$age, decreasing = TRUE)[1:5]
titanic[oldest, -3]
```

## Select the maxima or minima
```{r}
titanic[which.min(titanic$age), ]
titanic[which.max(titanic$age), ]
```

---
```{r}
titanic[which.max(titanic$age), ]

# gives the same result but harder to read
titanic[which(titanic$age == max(titanic$age)), ]
```






