---
title: "Introduction to R: Visualizations with `ggplot2`"
author: "Malte Bonart"
output: revealjs::revealjs_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, fig.height = 3)
```

# Basics

## Plotting systems in `R`
- `basegraphs`
- `grid` und `lattice`
- **`ggplot2`**

## Background
- system for declaratively creating graphics, based on the "*grammar of graphics*"
- special syntax: components are added using the `+` operator
- relevant data to plot should be collected in a `data.frame`
- produces high quality plots and takes care of many details

## Workflow

- initialize a plot with the `ggplot()` function
- supply a dataset 
- map *aesthetics* to variables: x, y, color, groups
- add layers/`geom`: points, lines, histogram
- add scales, faceting specification, coordinate systems, themes
- save the plot

## Installation of the ggplot2 package
```{r, echo = TRUE, eval = FALSE}
install.packages("ggplot2")
```

# Step by step guide

## The data
use the build in dataset `failthful` 
```{r}
?faithful
head(faithful)
```

## Initialize the plot
any variables that are part of the source dataframe have to be provided inside the `aes()` function

```{r}
library(ggplot2)
ggplot(faithful, aes(x = waiting, y = eruptions))
```

---

```{r, echo = FALSE}
library(ggplot2)
ggplot(faithful, aes(x = waiting, y = eruptions))
```


## Add some points
additional layers have to be "added" with the `+` operator
```{r}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
  geom_point()
```

---

```{r, echo = FALSE}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
  geom_point()
```

## Add a linear trend line
```{r}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
  geom_point() + 
  geom_smooth(method='lm')
```

---

```{r, echo = FALSE}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
  geom_point() + 
  geom_smooth(method='lm')
```


## Change the color and size
```{r}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
  geom_point(col ="steelblue", size = 0.9) + 
  geom_smooth(method = 'lm', color = "black")
```

---

```{r, echo = FALSE}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
  geom_point(col ="steelblue", size = 0.9) + 
  geom_smooth(method = 'lm', color = "black")
```

## Add some labels
```{r}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
 geom_point(col ="steelblue", size = 0.9) + 
  geom_smooth(method = 'lm', color = "black") + 
  labs(title = "Old Faithful Geyser Data", 
       subtitle = "Waiting time between eruptions and the duration of the eruption", 
       x = "waiting time in mins", y = "eruption time in mins"
       )
```

---

```{r, echo = FALSE}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
 geom_point(col ="steelblue", size = 0.9) + 
  geom_smooth(method = 'lm', color = "black") + 
  labs(title = "Old Faithful Geyser Data", 
       subtitle = "Waiting time between eruptions and the duration of the eruption", 
       x = "waiting time in mins", y = "eruption time in mins"
       )
```

## Change the theme
```{r}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
 geom_point(col ="steelblue", size = 0.9) + 
  geom_smooth(method = 'lm', color = "black") + 
  labs(title = "Old Faithful Geyser Data", 
       subtitle = "Waiting time between eruptions and the duration of the eruption", 
       x = "waiting time in mins", y = "eruption time in mins"
       ) + 
  theme_minimal()
```

---

```{r, echo = FALSE}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
 geom_point(col ="steelblue", size = 0.9) + 
  geom_smooth(method = 'lm', color = "black") + 
  labs(title = "Old Faithful Geyser Data", 
       subtitle = "Waiting time between eruptions and the duration of the eruption", 
       x = "waiting time in mins", y = "eruption time in mins"
       ) + 
  theme_minimal()
```

## Simple Linear Regression

```{r}
m <- lm(eruptions ~ waiting, data = faithful)
m <- summary(m)
m
rSquared <- round(m$r.squared, 2)
```

## Add R^2 to the plot
```{r}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
 geom_point(col ="steelblue", size = 0.9) + 
  geom_smooth(method = 'lm', color = "black") + 
  labs(title = "Old Faithful Geyser Data", 
       subtitle = "Waiting time between eruptions and the duration of the eruption", 
       x = "waiting time in mins", y = "eruption time in mins"
       ) + 
 geom_label(x = 90, y = 1.5, size = 4,
            label = paste("Bestimmtheitsmaß:", rSquared))

```

---

```{r, echo = FALSE}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
 geom_point(col ="steelblue", size = 0.9) + 
  geom_smooth(method = 'lm', color = "black") + 
  labs(title = "Old Faithful Geyser Data", 
       subtitle = "Waiting time between eruptions and the duration of the eruption", 
       x = "waiting time in mins", y = "eruption time in mins"
       ) + 
 geom_label(x = 90, y = 1.5, size = 4,
            label = paste("Bestimmtheitsmaß:", rSquared))
```

# Basic examples

## Read in the data

```{r}
titanic <- read.csv("./www/titanic.csv", stringsAsFactors = FALSE)
soccer <- read.csv2("./www/football.csv", stringsAsFactors = FALSE)
```

## Histogram with subplots

```{r, eval = TRUE, echo = TRUE}
ggplot(titanic, aes(x = age)) + 
  geom_histogram(binwidth = 5, na.rm = TRUE) +
  facet_grid(~ sex)
```

## Bar chart by groups
```{r, eval = TRUE, echo = TRUE}
ggplot(titanic, aes(x = pclass, y = age, fill = sex)) + 
  geom_bar(stat = "summary", fun.y = "mean", position = "dodge")
```

## Boxplot by groups
```{r}
titanic$pclass <- as.character(titanic$pclass)
ggplot(titanic, aes(y = age, x = pclass)) + 
  geom_boxplot()
```

## Visualize contingency tables
```{r}
ggplot(titanic, aes(y = sex, x = pclass)) + 
  geom_count()
```

## Count entries by year
```{r}
library(lubridate)
soccer$date <- as_date(soccer$date)
soccer$year <- year(soccer$date)
ggplot(soccer, aes(x = year)) + 
  geom_line(stat = "count")
```

# References

## Links

- [ggplot2 refernce](http://ggplot2.tidyverse.org/reference/)
- [ggplot2 cheat sheet](https://www.rstudio.org/links/data_visualization_cheat_sheet)



