---
title: "Introduction to R: Further topics"
author: "Malte Bonart"
output: revealjs::revealjs_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```


## How to continue?
- Two additional important data structures in `R`
    - [`list`](#lists)
    - [`matrix`](#matrix)
- Iteration over data structures with the [`lapply`- function](#lapply)
- Aggregation and grouping with the [`dplyr` package](#dplyr)
- Linear regression models with the [`lm` function](#lm)
- Time and datetime values in `R` with the `lubridate` package
- String manipulation with the `stringr` package


## Save datasets
```{r, eval = FALSE}
titanic <- read.csv(file = "./www/titanic.csv", stringsAsFactors = FALSE)

# text format
write.csv(titanic, file = "./www/titanic.csv") 

write.table(titanic, file = "./www/titanic.csv", 
            sep = ";", row.names = FALSE)


# binary 'R' format
saveRDS(titanic, file = "./www/titanic.rds") 

save(titanic, file = "./www/titanic.RData")
```


## Overview of important data strucutures

```{r, echo = F, results = 'asis'}
tab <- data.frame(Homogen = c("vector c()", "matrix matrix()"), Hetergogen = c("list list()", "data.frame data.frame()"))
rownames(tab) <- c("Eindimensional", "Zweidimensional")
knitr::kable(tab, format = "html")
```

# Lists {#lists}
## Introduction
- One dimensional data structure
- Each entry can be an arbitrary data strucutre (vector, matrice, data frame, function, list)
- "*Lists are like freight trains*"
- Usefull inside functions to return several results with different data structures

## Example
```{r}
l <- list(values = 1:10, fun = sum, innerlist = list(1, 2))
l
length(l)
```

## List element extraction: `L[...]`
- Result of an extraction is always a *list*
- Methods of extraction are the same as for `vectors` or `data.frames`
    - logical vector
    - positive (inclusive) or negative (exclusive) indices
    - character vector (if list has `names`)
    
---

```{r, eval = TRUE, echo = TRUE}
l[c(TRUE, FALSE, FALSE)]
l[c("innerlist", "fun")]
```


## Data extraction: `L[[...]]`
- Extracts the *content* inside *one* list element 
- Only *one* element can be extracted
- `$`-sign extraction possible (if list has `names`)

---

```{r, error=TRUE}
l[[1]]
l$values
l[["fun"]]
l[["fun"]](c(1, 2, 3)) # directly apply the 'sum' function
```

# Matrices {#matrix}
## Introduction
- Two-dimensional data structure where each entry has the same data-type
    - Like `vector` but two-dimensional
    - Like `data.frame` but only one data-type allowed (`numeric`, `logical`, `character`)
- Special operators and functions for matrix algebra
    - Inversion: `solve()`
    - Multiplication: `%*%`
    - Transpose: `t()`

## Creation
- `matrix()` function with at least two arguments:
    - a vector `c()` of elements
    - `nrow` or `ncol` specification
    - matrix is filled by columns unless `byrow = TRUE` 

## Example
```{r}
mat <- matrix(c(1, 3, 2, 6, 4, 5), ncol = 3)
mat
matrix(c(1, 3, 2, 6, 4, 5), ncol = 3, byrow = TRUE)
c(ncol(mat), nrow(mat))
```

## Extration `M[i, j]`
- Methods of extraction are the same as for `vectors` or `data.frames`
    - logical vector
    - positive (inclusive) or negative (exclusive) indices
    - character vector (if matrix has `rownames` or `colnames`)

```{r}
mat[c(1, 2), c(1, 2)]
```

# The lapply family {#lapply} 

## Idea
- `apply`: iteratation over the columns or rows of a `matrix`
- `lapply`, `sapply`: iteration over a `vector`, `list` or the columns of a `data.frame`
- avoids code repitions and errors
- code becomes easier to read
- alternative to `for` loops 

## The sapply function
`sapply(X, FUN, ...)`

- `X` is a list a vector or a data frame
- `FUN` is a function which is applied on each column of a data frame or each entry of the list or each value of a vector
- `...` additional arguments of the function specified in `FUN`

## Example I
```{r}
titanic <- read.csv("./www/titanic.csv", stringsAsFactors = FALSE)
countNA <- function(x){
  return(sum(is.na(x)))
}
sapply(titanic, countNA)
```

## Example II
```{r}
l <- list(rnorm(100), rnorm(100, mean = 3), rnorm(100, mean = -3))
lapply(l, mean)
```

## `for` loop
```{r}
result <- rep(NA, 100)
for(i in 1:100){
  result[i] <- mean(rnorm(n = 1000))
}
```

## `sapply` loop
```{r}
result <- sapply(1:100, function(i) {
  mean(rnorm(n = 1000))
})
```

# `dplyr` package: Grouping and aggregation of data frames {#dplyr}

## Introduction

- The `dplyr` package provides a powerfull *grammar* of data analysis
- Commands are chained together with the piping operator `%>%`
- Easier and cleaner way for a lot of common `data.frame` operations:
    - grouping, 
    - sorting
    - filtering
    - aggregation
    - conversion between *long* and *wide* format (with the `tidyr` package)
    
## Example: Grouping and aggregation
```{r}
library(dplyr)
titanic <- read.csv("./www/titanic.csv", stringsAsFactors = FALSE)
titanic %>%
  group_by(pclass, sex) %>%
  summarize(survived = mean(survived), meanAge = mean(age, na.rm = TRUE))
```

## Reference
- [reference](http://dplyr.tidyverse.org/)

# Linear regression models {#lm}

## Introduction
- Linear regression is used in many areas
- Model the linear relationship between an outcome $y$ and $k$ independent variables $x_j, j = 1, \dots, k$. 
- $y_i = \beta_0 + \beta_1 x_{i1} + \dots + \beta_k x_{ik} + \epsilon_i$
- $\epsilon \overset{iid}{\sim} N(0, \sigma^2)$ 


## Simple linear regression in R

```{r}
titanic <- read.csv("./www/titanic.csv", stringsAsFactors = FALSE)
model <- lm(survived ~ sex, data = titanic)
summary(model)
```

## Categorical variables
```{r}
titanic$pclass <- as.factor(titanic$pclass)
model <- lm(survived ~ pclass, data = titanic)
summary(model)
```

## Several variables
```{r}
titanic$embarked <- as.factor(titanic$embarked)
model <- lm(survived ~ pclass + embarked, data = titanic)
summary(model)
```

## Predictions
```{r}
pred <- predict.lm(model)
survived_pred <- pred > 0.5
table(titanic$survived, survived_pred)
```

## Out of sample predictions
```{r}
model <- lm(survived ~ pclass + sex, data = titanic)
new <- data.frame(pclass = c("1", "2"), sex = c("male", "male"))
predict.lm(model, newdata = new, interval = "prediction", level = 0.90)
```


